---
title: "Gender Differences in Australian Political Parties"
author: "Sabrina Kreyzerman"
thanks: "Code and data from this analysis are available at: https://github.com/sabrinakreyzerman-k/Donaldson_Paper_Gender_in_Australian_Elections."
date: today
date-format: long
abstract: "Gender disparities among political candidates, within and across political parties, are not novel. Using data sourced from the Australian Electoral Commission (AEC), this analysis highlights the stark difference in female and male political candidates in the Australian House of Representatives. In a positive light, this paper will also support that there have been recent advancements in the number of female Australian political candidates serving as Members of Parliament, for instance, through gender quotas within parties."
format:
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
bibliography: references.bib
---

```{r}
#| echo: false
#| message: false
#| warning: false

library(tidyverse) # A collection of data-related packages
library(janitor) # Helps clean datasets
library(AustralianPoliticians)

```

# Introduction

Introduction goes here.

# Methodology

## Data Collection

Data Collection & Methodology goes here.

# Findings

Findings goes here.

```{r}
#| echo: false
#| message: false
#| warning: false

#### Download data ####
politician_data <- get_auspol("all") 

party_data <- get_auspol("allbyparty")

mp_data <- get_auspol("mps")

#### Clean data ####
clean_politician_data <- politician_data |>
  select(uniqueID, displayName, gender, birthDate, member)

clean_party_data <- party_data |>
  select(uniqueID, partyAbbrev, partyName, partySimplifiedName)

clean_mp_data <- mp_data |>
  select(uniqueID, mpFrom, mpTo)

#### Save dataset ####
write_csv(x = clean_politician_data, file = "politicians_data.csv")

write_csv(x = clean_party_data, file = "party_data.csv")

write_csv(x = clean_mp_data, file = "mp_data.csv")

#### Match up datasets ####
 clean_politician_data <- clean_politician_data |>
  left_join(clean_party_data, by = "uniqueID")

clean_politician_data <- clean_politician_data |>
  left_join(clean_mp_data, by = "uniqueID")

#### Save match up dataset ####
write_csv(clean_politician_data, "combined_politician_dataset.csv")

#### Read in dataset ####
combined_dataset <- read_csv("combined_politician_dataset.csv")

#### Read in the data ####
## Code sourced from: https://tellingstorieswithdata.com/02-drinking_from_a_fire_hose.html#australian-elections
raw_elections_data <-
  read_csv(
    file = 
      "https://results.aec.gov.au/27966/website/Downloads/HouseMembersElectedDownload-27966.csv",
    show_col_types = FALSE,
    skip = 1)

# We have read the data from the AEC website. We may like to save
# it in case something happens or they move it.
write_csv(
  x = raw_elections_data,
  file = "australian_voting.csv"
)

```


```{r}
#| message: false
#| echo: false
#| warning: false

## Load packages ##
library(tidyverse)
library(knitr)
library(janitor)
library(kableExtra)

## Read in data ##
combined_dataset <- read_csv("combined_politician_dataset.csv")
```

For more information, see our @tbl-gender_party. 

```{r}
#| message: false
#| echo: false
#| label: tbl-gender_party
#| tbl-cap: Number of MPs broken down by gender and party

combined_dataset |> 
  filter(partySimplifiedName == "Liberals" | partySimplifiedName == "Labor") |>
  group_by(gender, partySimplifiedName) |>
  count() |>
  ungroup() |>
  mutate(percentage = round(n / sum(n) * 100, 2)) |>
  kable(
    col.names = c("Gender", "Party", "Number of MPs",	"Percentage of MPs"),
  booktabs = TRUE)
```

```{r}
#| echo: false
#| message: false
#| warning: false

simulated_data <-
  tibble(
    # Use 1 through to 151 to represent each division
    "Division" = 1:151,
    # Randomly pick an option, with replacement, 151 times
    "Party" = sample(
      x = c("Liberal", "Labor", "National", "Green", "Other"),
      size = 151,
      replace = TRUE
    )
  )

simulated_data

```

```{r}
#| echo: false
#| message: false
#| warning: false

#### Read in the data ####
raw_elections_data <-
  read_csv(
    file = 
      "https://results.aec.gov.au/27966/website/Downloads/HouseMembersElectedDownload-27966.csv",
    show_col_types = FALSE,
    skip = 1
  )

# We have read the data from the AEC website. We may like to save
# it in case something happens or they move it.
write_csv(
  x = raw_elections_data,
  file = "australian_voting.csv"
)

```

```{r}
#| echo: false
#| message: false
#| warning: false

#### Basic cleaning ####
raw_elections_data <-
  read_csv(
    file = "australian_voting.csv",
    show_col_types = FALSE
  )

```

```{r}
#| echo: false
#| message: false
#| warning: false

# Make the names easier to type
cleaned_elections_data <-
  clean_names(raw_elections_data)

```

```{r}
#| echo: false
#| message: false
#| warning: false

cleaned_elections_data <-
  cleaned_elections_data |>
  select(
    division_nm,
    party_nm
  )

```

```{r}
#| echo: false
#| message: false
#| warning: false

names(cleaned_elections_data)
```

```{r}
#| echo: false
#| message: false
#| warning: false

cleaned_elections_data <-
  cleaned_elections_data |>
  rename(
    division = division_nm,
    elected_party = party_nm
  )

```

```{r}
#| echo: false
#| message: false
#| warning: false

cleaned_elections_data$elected_party |>
  unique()
```

```{r}
#| echo: false
#| message: false
#| warning: false

cleaned_elections_data <-
  cleaned_elections_data |>
  mutate(
    elected_party =
      case_match(
        elected_party,
        "Australian Labor Party" ~ "Labor",
        "Liberal National Party of Queensland" ~ "Liberal",
        "Liberal" ~ "Liberal",
        "The Nationals" ~ "Nationals",
        "The Greens" ~ "Greens",
        "Independent" ~ "Other",
        "Katter's Australian Party (KAP)" ~ "Other",
        "Centre Alliance" ~ "Other"
      )
  )

head(cleaned_elections_data)
```

```{r}
#| echo: false
#| message: false
#| warning: false

write_csv(
  x = cleaned_elections_data,
  file = "cleaned_elections_data.csv"
)
```

```{r}
#| echo: false
#| message: false
#| warning: false

#### Read in the data ####
cleaned_elections_data <-
  read_csv(
    file = "cleaned_elections_data.csv",
    show_col_types = FALSE
  )
```

```{r}
#| echo: false
#| message: false
#| warning: false

cleaned_elections_data |>
  count(elected_party)
```

```{r}
#| echo: false
#| message: false
#| warning: false

cleaned_elections_data |>
  ggplot(aes(x = elected_party)) + # aes abbreviates "aesthetics" 
  geom_bar()

cleaned_elections_data |>
  ggplot(aes(x = elected_party)) +
  geom_bar() +
  theme_minimal() + # Make the theme neater
  labs(x = "Party", y = "Number of seats") # Make labels more meaningful

```
# Discussion

Discussion goes here.

# Exclusions & Future Research

One variable I chose to exclude from my data set was the age of political candidates. This decision was made because my objective with this analysis was to zero-in on the implications of gender within and across political parties. There is research to support that age is also at the forefront of considerations taken by voters when exercising their electoral rights (…). This is a compelling site for future study. While beyond the scope of my analysis, future research can look to advance … . Future research can focus on how age either increases or decreases chances of election, why voters may be deterred or compelled by younger or older candidates, and how age impacts electoral results domestically and cross-nationally.

# Conclusion

Conclusion goes here.

# References